---
title: "Spatial Machine Learning for Air Pollution Modeling in Poland"
subtitle: "Predicting Ground-Level NO₂ Using Sentinel-5P TROPOMI, Ground Stations & Spatial Features"
author: "Ondřej Marvan & Adam Jaworski"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
    fig_width: 10
    fig_height: 7
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.align = "center", cache = TRUE
)

source("00_setup.R")

library(sf)
library(terra)
library(tidyverse)
library(spdep)
library(ranger)
library(gstat)
library(automap)
library(dbscan)
library(cluster)
library(viridis)
library(corrplot)
library(knitr)
```

# 1. Introduction

Air pollution remains one of the most pressing public health challenges in
Poland, with nitrogen dioxide (NO₂) concentrations regularly exceeding EU
limits in major urban areas and along transport corridors. This study combines
satellite-based atmospheric measurements from the **Sentinel-5P TROPOMI**
instrument with ground-level observations from the **GIOŚ** national monitoring
network to build spatial machine learning models predicting NO₂ concentrations
across Poland at 5 km resolution.

**Research questions:**

1. How well can satellite-derived tropospheric NO₂ columns predict ground-level
   concentrations?
2. Which spatial features (road density, industrial land use, population)
   improve prediction accuracy?
3. Where are the persistent pollution hotspots, and what spatial patterns
   emerge?
4. Can kriging improve predictions by capturing residual spatial
   autocorrelation?


# 2. Data & Methods

```{r load-data}
grid     <- safe_read_sf(file.path(CONFIG$dir_processed, "analysis_grid.gpkg"))
tropomi  <- safe_read_csv(file.path(CONFIG$dir_processed, "tropomi_grid.csv"))
stations <- safe_read_sf(file.path(CONFIG$dir_processed, "stations_gridded.gpkg"))
```

## 2.1 Data Sources

| Source | Content | Resolution | Access method |
|--------|---------|-----------|---------------|
| Sentinel-5P TROPOMI | NO₂, SO₂, CO columns | ~5.5 km, daily → monthly | Google Earth Engine (`rgee`) |
| GIOŚ | Ground NO₂, SO₂, PM10, PM2.5, CO, O₃ | ~250 stations, hourly | REST API + bulk CSV |
| GUS BDL | Population, vehicles, industry | Gmina level | REST API |
| OpenStreetMap | Roads, railways, industrial zones | Vector | `osmdata` package |

## 2.2 Analysis Grid

We constructed a regular **`r CONFIG$grid_res / 1000` km** grid over Poland in
EPSG:2180 (Poland CS92), clipped to the national boundary.
Grid cells serve as the common spatial unit for integrating all data sources.

```{r grid-summary}
if (!is.null(grid)) {
  cat(sprintf("Grid cells: %d\n", nrow(grid)))
  cat(sprintf("Features: %s\n", paste(
    setdiff(names(grid), base::c("geometry", "cell_id", "x_coord", "y_coord")),
    collapse = ", "
  )))
}
```


# 3. Exploratory Data Analysis

## 3.1 NO₂ Spatial Distribution

```{r no2-map, fig.cap="Annual mean tropospheric NO₂ over Poland (TROPOMI)."}
if (!is.null(grid) && "tropomi_no2" %in% names(grid)) {
  ggplot(grid |> filter(!is.na(tropomi_no2))) +
    geom_sf(aes(fill = tropomi_no2), color = NA) +
    scale_fill_viridis(option = "inferno", name = expression(NO[2])) +
    labs(title = "Annual Mean Tropospheric NO₂ — Poland",
         subtitle = "Source: Sentinel-5P TROPOMI") +
    theme_minimal()
}
```

## 3.2 Monthly Temporal Patterns

```{r monthly-ts, fig.cap="Monthly national average pollutant concentrations."}
if (!is.null(tropomi)) {
  tropomi |>
    group_by(year_month, pollutant) |>
    summarise(mean_val = mean(value, na.rm = TRUE),
              sd_val   = sd(value, na.rm = TRUE), .groups = "drop") |>
    ggplot(aes(x = year_month, y = mean_val)) +
    geom_line(linewidth = 1, color = "steelblue") +
    geom_ribbon(aes(ymin = mean_val - sd_val, ymax = mean_val + sd_val),
                alpha = 0.2, fill = "steelblue") +
    facet_wrap(~pollutant, scales = "free_y", ncol = 1) +
    labs(x = NULL, y = "Column density (mol/m²)") +
    theme_minimal()
}
```

## 3.3 Feature Correlations

```{r correlation, fig.cap="Correlation matrix of grid features."}
if (!is.null(grid)) {
  num_cols <- grid |>
    st_drop_geometry() |>
    select(where(is.numeric), -cell_id, -x_coord, -y_coord)

  if (ncol(num_cols) >= 3) {
    cor_mat <- cor(drop_na(num_cols))
    corrplot(cor_mat, method = "color", type = "lower",
             tl.cex = 0.7, addCoef.col = "black", number.cex = 0.6)
  }
}
```

## 3.4 Spatial Autocorrelation

```{r moran}
if (!is.null(grid) && "tropomi_no2" %in% names(grid)) {
  clean <- grid |> filter(!is.na(tropomi_no2))
  coords <- st_coordinates(st_centroid(clean))
  nb <- knearneigh(coords, k = 8) |> knn2nb()
  lw <- nb2listw(nb, style = "W")
  moran_res <- moran.test(clean$tropomi_no2, lw)

  cat(sprintf("Moran's I = %.4f (p = %s)\n",
              moran_res$estimate[1], format.pval(moran_res$p.value)))
  cat("Interpretation: strong positive spatial autocorrelation confirms that",
      "NO₂ is spatially clustered, justifying spatial ML methods.\n")
}
```


# 4. Supervised Machine Learning

```{r supervised, results='hide'}
# Source the analysis functions
source("03_analysis.R")

# Run supervised ML (uncomment when data is ready)
# sup_results <- run_supervised()
```

## 4.1 Random Forest

```{r rf-importance, fig.cap="Top 15 predictors by impurity-based importance."}
imp_path <- file.path(CONFIG$dir_figures, "rf_importance.png")
if (file.exists(imp_path)) knitr::include_graphics(imp_path)
```

## 4.2 Geographically Weighted Random Forest

The GW-RF captures **spatial non-stationarity** — the idea that predictor
effects vary across space (e.g., road density matters more in urban corridors
than in rural eastern Poland).

## 4.3 Model Comparison

```{r model-comparison}
comp_path <- file.path(CONFIG$dir_output, "model_comparison.csv")
if (file.exists(comp_path)) {
  read_csv(comp_path, show_col_types = FALSE) |> kable(digits = 4)
}
```


# 5. Unsupervised Machine Learning

## 5.1 Hotspot Detection (DBSCAN)

```{r hotspots, fig.cap="DBSCAN clusters of top-quartile NO₂ grid cells."}
hot_path <- file.path(CONFIG$dir_figures, "hotspots_dbscan.png")
if (file.exists(hot_path)) knitr::include_graphics(hot_path)
```

## 5.2 LISA Cluster Map

```{r lisa, fig.cap="Local Moran's I — High-High clusters mark persistent pollution zones."}
lisa_path <- file.path(CONFIG$dir_figures, "lisa_no2.png")
if (file.exists(lisa_path)) knitr::include_graphics(lisa_path)
```

## 5.3 Spatial Typology (PAM Clustering)

```{r pam-clusters, fig.cap="Multi-variable spatial typology of Poland."}
pam_path <- file.path(CONFIG$dir_figures, "spatial_clusters.png")
if (file.exists(pam_path)) knitr::include_graphics(pam_path)
```


# 6. Kriging

## 6.1 Variogram

```{r variogram, fig.cap="Fitted semivariogram of ground-level NO₂."}
var_path <- file.path(CONFIG$dir_figures, "variogram.png")
if (file.exists(var_path)) knitr::include_graphics(var_path)
```

## 6.2 Interpolated Surface

```{r kriging-map, fig.cap="Ordinary Kriging prediction surface."}
ok_path <- file.path(CONFIG$dir_figures, "kriging_ok.png")
if (file.exists(ok_path)) knitr::include_graphics(ok_path)
```

## 6.3 Kriging Method Comparison

```{r kriging-comparison}
kc_path <- file.path(CONFIG$dir_output, "kriging_comparison.csv")
if (file.exists(kc_path)) {
  read_csv(kc_path, show_col_types = FALSE) |> kable(digits = 4)
}
```


# 7. Discussion

## Key Findings

<!-- Fill in after analysis -->

1. **TROPOMI–ground relationship**: ...
2. **Most important predictors**: ...
3. **Spatial non-stationarity**: GW-RF reveals where predictor effects vary
4. **Hotspot locations**: Persistent clusters in ...
5. **Kriging improvement**: Regression kriging captures residual spatial structure

## Limitations

- TROPOMI cloud masking reduces usable observations, especially in winter
- GIOŚ station network is denser in urban areas → potential spatial bias
- 5 km grid smooths very local gradients (street-level variation lost)
- Temporal mismatch: satellite overpass (~13:30) vs. 24h ground averages


# 8. Conclusion

<!-- Fill in after analysis -->


# References

- Copernicus Sentinel-5P (processed by ESA), TROPOMI L2 NO₂ products
- GIOŚ — Główny Inspektorat Ochrony Środowiska, Państwowy Monitoring Środowiska
- GUS — Bank Danych Lokalnych (BDL)
- OpenStreetMap contributors

---

*Spatial Machine Learning in R, WNE UW 2026 — prof. Katarzyna Kopczewska*

```{r session-info}
sessionInfo()
```
